### 2019-04-17 paper/response.txt

We are very grateful for the helpful comments that all reviews make, and are
especially gladened by the expressions of appreciation:

  * "extremely well-written"
  * "going significantly beyond [...] previous work"
  * "exemplary in its precision and rigor" 
  * " I enjoyed reading this paper"
  * ...

We will do our best to address all comments in the final version, but focus
here only on the main issues and questions raised.

@@

  We will articulate this better. You make a good point here about moving the concrete Dom 
  from the appendix to the main text. Then we will be able to constrast with ListT-done
  right applied to the state monad, which suffers from the above two problems.

- Review C:

  Novelty wrt "Just do It":
  TODO

  Jeremy Gibbons, one of the authors of "Just do It" acknolwedges our contributions as follows:

> The Just Do It paper deals only with the backtrackable combination of
> nondeterminism and state, where composition distributes rightwards over failure
> (ie failure is a right zero of composition) and over choice - your laws
> (15,16). We don’t deal with persistent state, your law (18). We didn’t know
> then what the right axiomatization was for persistent state, nor a suitable
> model for it - we confidently asserted (S7) that s->([a],s) is an alternative
> model of nondeterminism and state, but in fact I believe this is not even a
> monad. Your paper contributes a useful axiomatization, experimental validation
> of those axioms through a variant of N Queens that doesn’t copy the state, a
> model, and a Coq verification of that model.

@@

    As we require a particular interaction between state and non-determinism,
    namely the local state semantics, by definition we do not want to reason about
    the two effects in isolation.

    Reasoning about local state happens modularly, and not for a whole program 
    at once. We use the local state laws for that.

    Enabling this local reasoning about local state, when implemented by means of
    global state, requires a global discipline of modifying state only in a 
    restoring-upon-backtracking fashion. We carry out this proof once, which involves
    reasoning about all disciplined whole programs, so that we can then reason
    locally about all specific programs.

    (Algebraic) Effect Handlers Approach:
    "in the middle of the paper it is revealed that the reader should have read
     the paper interpreting certain things as syntax and certain as semantics. In
     addition the meaning (semantics) of = is described in vague terms informally in
     a cryptic note reference to event handlers"

    We do not intend for the the reader to have read the first part of the
    paper in this way. On the contratry, like the "Just do It" paper, we advocate
    high-level axiomatic reasoning based on algebraic laws to understand programs. 
    However, as an auxiliary proof device, briefly sketched in the note on p.8, and
    elaborated in more detail in Sections 6-7 (and in full detail in our Coq 
    mechanisation), we can reason about monadic programs by separating their
    synax and semantics. Specific notions of equality are given on lines 942 and
    979.

    The second part of the paper presents a proof technique,
    where 
    (Note that we never mention event handlers, which are quite different from effect handlers.)

### 2019-04-17 paper/response.txt

  2. The state at the end of a branch is transferred to (the beginning of)
     the next branch.

     => Without a bind, we know where a branch is at its end and can build in
        the transfer of the state to the next branch. With a bind, we would
        not know when we are at the end of the branch and would have to defer
        the transfer, leaving room for the possibility that an entirely different
        state value is passed to the next branch.:w

We will articulate this better. You make a good point here about moving the concrete Dom 
from the appendix to the main text. Then we will be able to constrast with ListT-done
right applied to the state monad, which suffers from the above two problems.

# Review C:
ur paper contributes a useful axiomatization, experimental validation
> of those axioms through a variant of N Queens that doesn’t copy the state, a
> model, and a Coq verification of that model.

## Novelty wrt "Just do It":

Jeremy Gibbons, "Just do It" author, acknowledges our contributions as follows:

@@

> The Just Do It paper deals only with the backtrackable combination of
> nondeterminism and state, where composition distributes rightwards over failure
> (ie failure is a right zero of composition) and over choice - your laws
> (15,16). We don’t deal with persistent state, your law (18). We didn’t know
> then what the right axiomatization was for persistent state, nor a suitable
> model for it - we confidently asserted (S7) that s->([a],s) is an alternative
> model of nondeterminism and state, but in fact I believe this is not even a
> monad. Your paper contributes a useful axiomatization, experimental validation
> of those axioms through a variant of N Queens that doesn’t copy the state, a
> model, and a Coq verification of that model.

## The use of Haskell:
We feel that the use of Haskell for programming and reasoning with monads is justified given
the extensive existing body of literature that does just that. Nevertheless,
we appreciate your pointing out that the accessibility of the paper will be improved when we 
add more explanation of concepts like hylomorphisms and scanl.

## Benefits of Monads:

A big benefit of monads is that it provides a pureful functional model of
effects. This has become an actual practice in Haskell and is a good model for
what happens in languages with algebraic effects & handlers. Yet, we should
emphasize that the focus of our paper is reasoning about monadic programs, not
to justify the use of monads. We are not aware of any other formalism used
to tackle the same problem for other effect models. Even if it would exist,
we believe that our work is still relevant because it directly applies to 
the above settings where monads are used.

## (Non-)Modularity:

There is modularity and non-modularity at different levels, which should not
be mixed up.

 -  As we require a particular interaction between state and non-determinism,
    namely the local state semantics, by definition we do not want to reason about
    the two effects in isolation.

 -  Reasoning about local state happens modularly, and not for a whole program 
    at once. We use the local state laws for that.

 -  Enabling this local reasoning about local state, when implemented by means of
    global state, requires a global discipline of modifying state only in a 
    restoring-upon-backtracking fashion. We carry out this proof once, which involves
    reasoning about all disciplined whole programs, so that we can then reason
    locally about all specific programs.

## (Algebraic) Effect Handlers Approach:

> "in the middle of the paper it is revealed that the reader should have read
>  the paper interpreting certain things as syntax and certain as semantics. In
>  addition the meaning (semantics) of = is described in vague terms informally in
>  a cryptic note reference to event handlers"

@@

We do not intend for the the reader to have read the first part of the
paper in this way. On the contratry, like the "Just do It" paper, we advocate
high-level axiomatic reasoning based on algebraic laws to understand programs. 
However, as an auxiliary proof device, briefly sketched in the note on p.8, and
elaborated in more detail in Sections 6-7 (and in full detail in our Coq 
mechanisation), we can reason about monadic programs by separating their
synax and semantics. Specific notions of equality are given on lines 942 and
979.

# Review D:

> It seems that the contributions of this paper (lines 43-51) lie in justifying
> these techniques by equational reasoning on monadic programs. I find these
> contributions relatively thin, because rather than deriving theorems from a set
> of standalone equational laws, the paper freely drops down to the level of
> free-monad syntax trees (lines 368-381).

We don't think this characterisation does justice to our contributions:

- ...

- ...

- Proving the correctness of global state with respect to local state has not
  been a triviality, but a substantial effort, not just to find the appropriate
  auxiliary lemmas, but to create the appropriate proof approach.

  We are not aware of any prior proof for this problem, mechanised or otherwise, and
  certainly not for the monadic setting that supports axiomatic reasoning about
  program equivalence. In the absence of any other approach that has been demonstrated to
  solve the problem, we do not understand what fault you see in our approach.

  Moreover, we believe that having a working approach that can be rused for
  similar problems will be highly relevant for reasoning about algebraic effects
  and handlers (see Local Algebraic Theories in Related Work), which are
  becoming increasinlgy more prominent and where programmers are writing algebras
  for free-monad syntax trees all the time.

# A1:

### 2019-04-17 paper/response.txt

We are very grateful for the helpful comments that all reviews make, and are
especially gladdened by the expressions of appreciation:

@@

We will do our best to address all comments in the final version, but focus
here only on the main issues and questions raised.

# Review B:

## About Dom not having a bind:

There are two key ideas that make the emulation of local state by means
of global state work:

@@

     => Without a bind, a carefully constructed disciplined Dom value cannot
        be extended in a sequential (i.e., bind-) fashion with a wild put.

  2. The state at the end of a branch is transferred to (the beginning of)
     the next branch.

     => Without a bind, we know where a branch is at its end and can build in
        the transfer of the state to the next branch. With a bind, we would
        not know when we are at the end of the branch and would have to defer
        the transfer, leaving room for the possibility that an entirely different
        state value is passed to the next branch.:w

We will articulate this better. You make a good point here about moving the concrete Dom
from the appendix to the main text. Then we will be able to contrast with ListT-done
right applied to the state monad, which suffers from the above two problems.

@@

> The Just Do It paper deals only with the backtrackable combination of
> nondeterminism and state, where composition distributes rightwards over failure
> (ie failure is a right zero of composition) and over choice - your laws
> (15,16). We don’t deal with persistent state, your law (18). We didn’t know
> then what the right axiomatization was for persistent state, nor a suitable
> model for it - we confidently asserted (S7) that s->([a],s) is an alternative
> model of nondeterminism and state, but in fact I believe this is not even a
> monad. Your paper contributes a useful axiomatization, experimental validation
> of those axioms through a variant of N Queens that doesn’t copy the state, a
> model, and a Coq verification of that model.

## The use of Haskell:
We feel that the use of Haskell for programming and reasoning with monads is justified given
the extensive existing body of literature that does just that. Nevertheless,
we appreciate your pointing out that the accessibility of the paper will be improved when we
add more explanation of concepts like hylomorphisms and scanl.

## Benefits of Monads:

A big benefit of monads is that it provides a purely functional model of
effects. This has become an actual practice in Haskell and is a good model for
what happens in languages with algebraic effects & handlers. Yet, we should
emphasize that the focus of our paper is reasoning about monadic programs, not
to justify the use of monads. We are not aware of any other formalism used
to tackle the same problem for other effect models. Even if it would exist,
we believe that our work is still relevant because it directly applies to
the above settings where monads are used.

@@

 -  Enabling this local reasoning about local state, when implemented by means of
    global state, requires a global discipline of modifying state only in a
    restoring-upon-backtracking fashion. We carry out this proof once, which involves
    reasoning about all disciplined whole programs, so that we can then reason
    locally about all specific programs.

## (Algebraic) Effect Handlers Approach:

> "in the middle of the paper it is revealed that the reader should have read
>  the paper interpreting certain things as syntax and certain as semantics. In
>  addition the meaning (semantics) of = is described in vague terms informally in
>  a cryptic note reference to event handlers"

We do not intend for the the reader to have read the first part of the
paper in this way. On the contrary, like the "Just do It" paper, we advocate
high-level axiomatic reasoning based on algebraic laws to understand programs.
However, as an auxiliary proof device, briefly sketched in the note on p.8, and
elaborated in more detail in Sections 6-7 (and in full detail in our Coq
mechanisation), we can reason about monadic programs by separating their
syntax and semantics. Specific notions of equality are given on lines 942 and
979.

@@

- ...

- Proving the correctness of global state with respect to local state has not
  been a triviality, but a substantial effort, not just to find the appropriate
  auxiliary lemmas, but to create the appropriate proof approach.

  We are not aware of any prior proof for this problem, mechanised or otherwise, and
  certainly not for the monadic setting that supports axiomatic reasoning about
  program equivalence. In the absence of any other approach that has been demonstrated to
  solve the problem, we do not understand what fault you see in our approach.

  Moreover, we believe that having a working approach that can be used for
  similar problems will be highly relevant for reasoning about algebraic effects
  and handlers (see Local Algebraic Theories in Related Work), which are
  becoming increasingly more prominent and where programmers are writing algebras
  for free-monad syntax trees all the time.

### 2019-04-17 paper/response.txt

- Proving the correctness of global state with respect to local state has not
  been a triviality, but a substantial effort, not just to find the appropriate
  auxiliary lemmas, but to create the appropriate proof approach.

  We are not aware of any prior proof for this problem, mechanised or otherwise, and
  certainly not for the monadic setting that supports axiomatic reasoning about
  program equivalence. In the absence of any other approach that has been demonstrated to
  solve the problem, we do not understand what fault you see in our approach.

  Moreover, we believe that having a working approach that can be used for
  similar problems will be highly relevant for reasoning about algebraic effects
  and handlers (see Local Algebraic Theories in Related Work), which are
  becoming increasingly more prominent and where programmers are writing algebras
  for free-monad syntax trees all the time.

>  I have serious doubts about the first side condition on line 491, proven on line 543.

  That is indeed a mistake -- we copy-and-pasted an earlier, wrong version of the lemma. Lemma 3 should come with a side-condition: m commutes with state, while the "n" in Lemma 2 should be unfoldM p f. In our particular application, unfoldM p f = perm, which does satisfy the side conditions because non-determinism is its only effect.

> The assumptions that "functions are total" and "lists are inductive types" (line 96) are apparently violated

  Indeed zipWith uses infinite lists, and ok is partial. These can be easily fixed at the cost of losing some clarity. We (and we think, the readers) would prefer to keep it this way. [SCM: I don't think this is worth responding and can/should be omitted, given that we are running out of space.]

### 2019-04-18 paper/response.txt

  We are not aware of any prior proof for this problem, mechanised or otherwise, and
  certainly not for the monadic setting that supports axiomatic reasoning about
  program equivalence. In the absence of any other approach that has been demonstrated to
  solve the problem, we do not understand what fault you see in our approach.

  Moreover, we believe that having a working approach that can be used for
  similar problems will be highly relevant for reasoning about algebraic effects
  and handlers (see Local Algebraic Theories in Related Work), which are
  becoming increasingly more prominent and where programmers are writing algebras
  for free-monad syntax trees all the time.

>  I have serious doubts about the first side condition on line 491, proven on line 543.

  That is indeed a mistake -- we copy-and-pasted an earlier, wrong version of the lemma. Lemma 5.2 should come with a side-condition: m commutes with state, while the "n" in Theorem 5.1 should be unfoldM p f z. In our particular application, unfoldM p f = perm, which does satisfy the side conditions because non-determinism is its only effect. We have attached the correct theorems and their proofs as an attachment.

### 2019-04-18 paper/response.txt

We are grateful for your helpful comments, and are especially gladdened your
appreciation:

  * "extremely well-written"
  * "going significantly beyond [...] previous work"
  * "exemplary in its precision and rigor"
  * "I enjoyed reading this paper"
  * ...

We will do our best to address all comments in the final version, but focus
here only on the main ones.

@@

# Review B:

## About Dom not having a bind:

There are two key ideas that make the emulation of local state by means
of global state work:

  1. We impose a discipline that all uses of put are state-restoring;
     no `wild' puts are allowed.

     => Without a bind, a disciplined Dom value cannot be extended in a
        sequential (i.e., bind-) fashion with a wild put.

@@

  2. The state at the end of a branch is transferred to (the beginning of)
     the next branch.

     => Without a bind, we know where a branch is at its end and where to transfer
        the state to the next branch. With a bind, we would not know when we are at the
        end of the branch and would have to defer the transfer, allowing for the
        possibility that an entirely different state value is passed to the next
        branch.

We will articulate this better, incorporating your suggestion about moving the concrete Dom
from the appendix to the main text.

# Review C:

## Background

Given the extensive body of work on programming and reasoning with monads, we
believe it is justified to rely on some background from the reader.
Nevertheless, we hear your request to improve the accessibility by explaining
concepts like hylomorphisms and scanl in more detail.

@@

## (Algebraic) Effect Handlers Approach:

> "in the middle of the paper it is revealed that the reader should have read
>  the paper interpreting certain things as syntax and certain as semantics. In
>  addition the meaning (semantics) of = is described in vague terms informally in
>  a cryptic note reference to event handlers"

This is not what we intended. On the contrary, like the "Just do It" paper, we
advocate high-level axiomatic reasoning based on algebraic laws to understand
programs. However, as an auxiliary proof device, briefly sketched on p.8, and
elaborated in more detail in Sections 6-7 (and in our Coq mechanisation), we
can reason about monadic programs by separating their syntax and semantics.
Specific notions of equality are given on lines 942 and 979.

## Implications of a Domain without Bind

There are no implications, and that's precisely the point. The local state
axioms act like an interface based on which we can reason, and which decouples
the reasoning from any particular implementation. Sections 6-7 show that the
global state domain provides a valid implementation of the 'interface'.
The beauty of having reasoned against the interface is that we can use the
implementation without having to revisit any of the earlier proofs.

## (Non-)Modularity / Encapsulation:

 - As explained above, the axioms decouple the reasoning from a particular
   implementation and thus provide modularity.

 -  As we require a particular interaction between state and non-determinism,
    namely the local state semantics, by definition we do not want to reason about
    them in isolation.

 -  Satisfying the local state axioms by means of global state, requires a
    global discipline of only state-restoring updates.
    We carry out this proof once for all disciplined global state programs.
    The proof is then encapsulated in the local state axioms, so that
    anyone can reason a locally about the equivalence of specific programs
    based on those axioms.

In a personal communication, Jeremy Gibbons, "Just do It" author, states the
novelty as follows:

@@

> The Just Do It paper deals only with the backtrackable combination of
> nondeterminism and state, where composition distributes rightwards over failure
> (ie failure is a right zero of composition) and over choice - your laws
> (15,16). We don’t deal with persistent state, your law (18). We didn’t know
> then what the right axiomatization was for persistent state, nor a suitable
> model for it - we confidently asserted (S7) that s->([a],s) is an alternative
> model of nondeterminism and state, but in fact I believe this is not even a
> monad. Your paper contributes a useful axiomatization, experimental validation
> of those axioms through a variant of N Queens that doesn’t copy the state, a
> model, and a Coq verification of that model.

# Review D:

> It seems that the contributions of this paper (lines 43-51) lie in justifying
> these techniques by equational reasoning on monadic programs. I find these
> contributions relatively thin, because rather than deriving theorems from a set
> of standalone equational laws, the paper freely drops down to the level of
> free-monad syntax trees (lines 368-381).

We feel that this does not do justice to our contributions:

- The "Novelty wrt "Just do It"" paragraph above enumerates a range of novelties
  compared to the earlier work.

@@

- Proving the correctness of global state with respect to local state has not
  been a triviality, but a substantial effort, not just to find the appropriate
  auxiliary lemmas, but to create the appropriate proof approach.

- We are not aware of any prior proof for this problem, mechanised or
  otherwise, for the monadic setting (or any other) that supports
  axiomatic reasoning about program equivalence. In the absence of any other
  approach that has been demonstrated to solve the problem, we do not understand
  what fault you see in our approach.

- We believe that our approach is highly relevant because it can be used for 
  similar problems when reasoning about algebraic effects and handlers (see Local
  Algebraic Theories in Related Work), which are becoming increasingly more
  prominent and where programmers are writing algebras for free-monad syntax
  trees all the time.

> I have serious doubts about the first side condition on line 491, proven on line 543.

That is indeed a mistake -- we inadvertently copied an earlier, wrong
version of the lemma into the text. Lemma 5.2 should come with a
side-condition: m commutes with state, while the "n" in Theorem 5.1 should be
unfoldM p f z. In our particular application, unfoldM p f = perm, which does
satisfy the side conditions because non-determinism is its only effect. We have
included the correct theorems and their proofs in attachment.
